---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library(dplyr)
act_df <- read.csv("activity.csv", header = TRUE, na.strings = "NA") 
act_df$date <- as.POSIXct(act_df$date)
summary(act_df)
```

## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
per_day <- act_df %>%
    tbl_df() %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarize(all_steps = sum(steps))
mean_per_day <- mean(per_day$all_steps)
median_per_day <- median(per_day$all_steps)
p <- ggplot(per_day, aes(date, all_steps)) + geom_histogram(stat="identity")
print(p)
```

The average number of steps per day is `r mean_per_day` and the median of the same is `r median_per_day`

## What is the average daily activity pattern?

```{r}
per_interval <- act_df %>%
    tbl_df() %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>%
    summarize(ave_steps = mean(steps))
ave_ts = ts(per_interval$ave_steps)
plot.ts(ave_ts)
max_ave_steps <- max(per_interval$ave_steps)
max_interval_idx <- which.max(per_interval$ave_steps)
max_interval <- per_interval$interval[max_interval_idx]
```
The maximum average number of steps taken in a five minute interval is `r max_ave_steps` during the interval `r max_interval_idx`

## Imputing missing values

```{r}
# The strategy to replace the missing NA values is to: 
#  - calculate the mean of the average steps per interval
#  - replace the NA values with the mean calculated above
mean_interval_steps <- mean(per_interval$ave_steps)
imputed_act_df <- act_df %>% 
    mutate(imputed_steps = 
               ifelse(is.na(steps), 
                      mean_interval_steps, 
                      steps))
# Since the next section (weekend vs weekday) will use the resultant 
# data frame, I saved the data frame with the replaced NA first before
# calculating the mean per day in dplyr
imputed_mean_per_day <- imputed_act_df %>%
    tbl_df() %>%
    group_by(date) %>%
    summarize(all_steps = sum(imputed_steps))
imp_mean_per_day <- mean(imputed_mean_per_day$all_steps)
imputed_median_per_day <- median(imputed_mean_per_day$all_steps)
p <- ggplot(imputed_mean_per_day, aes(date, all_steps)) + geom_histogram(stat="identity")
print(p)
```

The average number of steps per day is `r imp_mean_per_day` and the median of the same is `r imputed_median_per_day`

The interesting part of this result is when the NA values are replaced with the average number of steps per interval, the mean value is unchanged, but the median value is now equal to the mean.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
daytype_act_df <- imputed_act_df %>%
    tbl_df() %>%
    mutate(day_type = ifelse(weekdays(date) == "Sunday" | weekdays(date) == "Saturday", "weekend", "weekday")) %>%
    group_by(interval, day_type) %>%
    summarise(all_steps = sum(imputed_steps))
par(mfrow=c(2,1))
weekend_ts = ts(daytype_act_df$all_steps[daytype_act_df$day_type=="weekend"])
weekday_ts = ts(daytype_act_df$all_steps[daytype_act_df$day_type=="weekday"])
plot.ts(weekend_ts)
plot.ts(weekday_ts)
```
